\section{Evaluation}
\label{sec:evaluation}

The evaluation is organized around the following reproducible experiments.
Result numbers are not filled in until the corresponding experiment is run with
fixed agent versions, logs, and statistical analysis.

\subsection{Research Questions}

The evaluation targets six questions.

\begin{description}
\item[RQ1: Bypass coverage.] Does OS-level enforcement catch the same policy
violation through direct tool calls, shell strings, subprocesses, SDK paths, and
direct syscalls?
\item[RQ2: Corrective feedback.] Does semantic feedback improve task completion
and reduce repeated violations compared with failure without a reason?
\item[RQ3: False positives.] Do benign tasks and sanctioned gate/declassify
paths proceed without unnecessary failure?
\item[RQ4: Overhead.] What are the per-event and end-to-end costs of label
propagation and rule checks?
\item[RQ5: Real-policy coverage.] How many real instructions from the corpus
are observable, expressible, robust, and temptable?
\item[RQ6: Baselines.] Which capabilities are unique relative to tool-layer
agent guardrails and kernel provenance/enforcement systems?
\end{description}

\subsection{Experiments}

\begin{table*}[t]
\centering
\caption{Planned evaluation matrix. Result fields remain \textsc{TBD}.}
\Description{A table listing planned experiments and planned result fields.}
\begin{tabular}{p{0.12\textwidth}p{0.26\textwidth}p{0.26\textwidth}p{0.24\textwidth}}
\toprule
Experiment & Method & Primary metrics & Result status \\
\midrule
E1: Cross-path coverage &
Run the same forbidden action through tool call, nested shell, subprocess, SDK/helper, and direct syscall triggers. Compare against tool-layer baselines. &
Coverage matrix over policy by path; feedback delivery. &
\textsc{TBD} \\
\midrule
E2: Feedback loop &
Run Codex and Claude-style agents under prompt-only, audit, fail-without-reason, and fail-with-feedback conditions. &
Task completion, repeated violations, repair iterations, token/turn cost. &
\textsc{TBD} \\
\midrule
E3: False positives &
Run allowed tasks and sanctioned paths: redaction, test-before-commit, confirmation, and lineage mediation. &
False positive rate, allowed-path success. &
\textsc{TBD} \\
\midrule
E4: Overhead &
Microbench fork/exec/read/write/connect and macrobench real agent tasks with and without ActPlane. &
p50/p99 latency, wall time, memory/map usage, rule-table scaling. &
\textsc{TBD} \\
\midrule
E5: Corpus coverage &
Double-code real instruction-file rules with D1--D7, report agreement, and derive policies/tasks mechanically. &
Observable, expressible, robust, temptable fractions; Cohen's kappa. &
\textsc{TBD} \\
\midrule
E6: Baseline comparison &
Compare against CamQuery, Tetragon, Progent, AgentSpec, and built-in agent sandboxes where runnable. &
Layer, enforce/detect, cross-channel taint, bypass resistance, feedback. &
\textsc{TBD} \\
\bottomrule
\end{tabular}
\label{tab:eval-plan}
\end{table*}

\subsection{Feedback Conditions}

The agent-facing experiment uses four conditions to isolate the contribution of
semantic feedback.

\begin{itemize}
\item C1: prompt-only. The natural-language rule is present, but there is no
ActPlane enforcement.
\item C2: audit. ActPlane observes and reports, but does not fail the action.
\item C3: fail without ActPlane reason. The violating action fails through
\texttt{-EPERM} or \texttt{SIGKILL}, but the agent receives no structured
remediation.
\item C4: fail with feedback. The action fails and ActPlane injects the rule
reason and remediation through the feedback file and hook bridge.
\end{itemize}

The main comparison is C3 to C4: the net benefit of the corrective feedback
loop. C1 to C2 estimates how often the prompt-only harness fails; C2 to C3
separates observation from action failure.

\subsection{Artifact Expectations}

The artifact should include policies, trigger scripts, agent prompts, raw logs,
statistical notebooks, and a container or VM description pinning kernel, BPF,
agent, and model versions. The final paper should avoid reporting any number
that cannot be regenerated from these artifacts.


\section{Related Work}
\label{sec:related}

\subsection{In-Kernel Provenance and Information Flow}

ActPlane builds on a long line of provenance and information-flow systems.
PASS made provenance a storage-layer primitive~\cite{muniswamy2006pass}.
Hi-Fi and LPM used LSM hooks to capture trustworthy system
provenance~\cite{pohly2012hifi,bates2015lpm}. CamFlow captured whole-system
provenance streams over processes, files, and sockets~\cite{pasquier2017camflow}.

CamQuery is the closest prior system~\cite{pasquier2018camquery}. It executes
provenance queries inline in the kernel, propagates labels across process, file,
and network edges, and can prevent policy violations. ActPlane does not claim
that this mechanism is new. The difference is the domain and substrate:
ActPlane targets cooperative AI agents, compiles an agent-oriented source/sink
DSL to eBPF/BPF-LSM, and returns corrective feedback to the actor that violated
the rule.

TaintDroid, Dytan, libdft, and Panorama established dynamic taint tracking at
runtime, binary-instrumentation, or emulation layers~\cite{enck2010taintdroid,
clause2007dytan,kemerlis2012libdft,yin2007panorama}. ActPlane uses a much
coarser object-level taint. That coarseness is what allows kernel-wide coverage,
but it creates an over-taint risk that must be measured.

\subsection{eBPF and LSM Enforcement}

BPF-LSM, originally proposed through KRSI, lets eBPF programs attach to Linux
security hooks and return allow/deny verdicts~\cite{singh2019krsi,linuxbpflsm}.
ActPlane uses this as its pre-operation backend. Contemporary eBPF security
tools and proposals such as Tetragon, Tracee, and eBPF-PATROL demonstrate
practical event capture and enforcement, but generally evaluate per-event
predicates or lineage flags rather than typed cross-channel information
flow~\cite{ciliumtetragon,aquasecuritytracee,ghimire2025ebpfpatrol}.

\subsection{Agent Guardrails}

AgentSpec and Progent are the closest agent-policy systems at the tool
layer~\cite{wang2025agentspec,shi2025progent}. They provide deterministic
checks over tool names, arguments, or framework actions. ActPlane complements these systems by
moving the enforcement point below the tool API. If all effects pass through the
framework, tool-layer policies are sufficient and cheaper. If the agent can
execute arbitrary code, spawn subprocesses, or use direct system interfaces, the
OS layer is the robust boundary.

FIDES and CaMeL show that typed information-flow and capability separation are
useful abstractions for LLM agents~\cite{costa2025fides,debenedetti2025camel}. They operate
inside the agent loop or scaffolding. ActPlane applies the same style of
information-flow reasoning at the OS boundary.

\subsection{Agent Instruction Files}

Recent empirical work studies \texttt{CLAUDE.md}, \texttt{AGENTS.md}, and other
agent context files: their structure, content, maintenance, and effect on task
efficiency~\cite{chatlatanagulchai2025manifests,chatlatanagulchai2025agentreadmes,
santos2025claudeconfig,lulla2026agentsmd}. ActPlane's corpus study asks a
different question: which instructions are behavioral or flow invariants, and
which of those can be observed and enforced below the tool layer?


\section{Discussion}

\subsection{Security Tool or Harness?}

ActPlane can express security-relevant policies, such as keeping secret-derived
data local. However, its primary target is an agent harness: a deterministic
operating contract for a cooperative but forgetful agent. This affects design
choices. A classic reference monitor would prioritize pre-operation denial for
all security claims. ActPlane also needs audit-only guidance, harness-level
termination when pre-operation arguments are unavailable, and remediation text
that lets the agent complete the original task through an allowed path.

\subsection{Where the Claim Must Stay Honest}

The strongest novelty claim is not the taint algorithm. In-kernel label
propagation and provenance enforcement are established ideas. ActPlane must be
evaluated as a systems contribution that combines those ideas with a new
workload and interface: autonomous agents with rich tool access, natural-language
policy sources, and corrective feedback.

The final paper should be careful about three boundaries.

\textbf{Block is LSM-only.} Tracepoint observation cannot be described as
\texttt{block}. The current implementation correctly ignores \texttt{block}
rules in tracepoint mode and warns the operator.

\textbf{Prompt files are aspirational.} A corpus of instruction files shows what
developers want agents to do. It does not by itself prove how often agents
violate those rules. That is the job of the agent evaluation.

\textbf{Coarse taint can over-approximate.} Object-level labels are practical,
but they may carry more provenance than a byte-level system would. The evaluation
must measure false positives and demonstrate useful declassification and gate
patterns.

\section{Conclusion}

ActPlane moves agent harness enforcement below the tool layer. It compiles
agent-oriented information-flow policies into eBPF/BPF-LSM checks, tracks labels
across process, file, and endpoint objects, and returns kernel-detected
violations as corrective feedback to the agent. The evaluation plan is designed
to test bypass coverage, feedback-driven recovery, false positives, overhead,
and coverage over real agent instruction files without reporting numbers before
they can be reproduced.
