\documentclass[sigplan,10pt]{acmart}

\setcopyright{none}
\renewcommand\footnotetextcopyrightpermission[1]{}
\settopmatter{printfolios=true}
\pagestyle{plain}

\acmConference[ATC '26]{2026 ACM SIGOPS Annual Technical Conference}{November 15--18, 2026}{Hyatt Hotel, Shatin, Hong Kong}
\acmYear{2026}
\acmISBN{}
\acmDOI{}

\title[ActPlane]{ActPlane: OS-Enforced Agent Harnesses}

\author{Paper \#XXX}
\affiliation{%
  \institution{Anonymous Submission}
  \city{}
  \country{}}

\begin{abstract}
AI coding agents are guided by natural-language instruction files that express
behavioral contracts: do not leak secrets, run tests before committing, use
only sanctioned tools. These contracts are declared at the level of
\emph{intent}, what the agent should and should not do, but their effects
occur as \emph{behavior}: actual OS operations such as file access, process
execution, and network connections. Between the two sits the agent's
\emph{action} layer: tool calls that map intent to OS behavior. However, existing agent harness enforcement operates at a single level, 
leaving an \emph{intent--behavior gap}:
intent-level prompts are forgettable and probabilistic;
action-level guards check tool calls but miss effects that reach the OS through
subprocesses or direct system interfaces; behavior-level enforcement (syscall
filters, sandboxes) is unbypassable but returns failures disconnected
from intent. A corpus study of popular projects shows that real contracts
frequently involve cross-object information flow in the OS, yet no existing harness
enforces them across intent and behavior.

ActPlane is the OS-level enforcement component of an agent harness. Agents and
developers declare behavioral contracts in a labeled information flow DSL; 
ActPlane compiles
them into rules enforced by an eBPF/BPF-LSM backend. Named labels propagate 
across process, file, and network objects at kernel
hooks, tracking which sources have influenced which subjects. When a violation
is detected, ActPlane returns corrective feedback to the agent: which contract
was violated, why, and what alternative path satisfies it.
We evaluate ActPlane on bypass coverage across five tool paths, agent recovery
rate under four feedback conditions, false positives, and per-event overhead.
\end{abstract}

\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>10010520.10010553.10010562</concept_id>
  <concept_desc>Computer systems organization~Embedded and cyber-physical systems</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>10002978.10003022.10003023</concept_id>
  <concept_desc>Security and privacy~Software and application security</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>10011007.10010940.10010941.10010949</concept_id>
  <concept_desc>Software and its engineering~Software safety</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[300]{Computer systems organization~Embedded and cyber-physical systems}
\ccsdesc[300]{Security and privacy~Software and application security}
\ccsdesc[300]{Software and its engineering~Software safety}

\keywords{AI agents, eBPF, BPF-LSM, labeled information-flow control, information-flow control, provenance, corrective feedback}

\begin{document}
\maketitle

\input{sections/01-introduction}
\input{sections/02-motivation}
\input{sections/03-design}
\input{sections/04-evaluation}

\bibliographystyle{ACM-Reference-Format}
\bibliography{references}

\end{document}
